---
title: "West Creek Floodplain Jam Analysis"
author: "Update from Juli"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  fontsize: 11pt
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pander)
library(kableExtra)
library(knitr)
library(Hmisc)
library(tibble)
library(car)
library(leaps)
library(MuMIn)
library(ggthemes)
library(tidyverse)

```

# Preliminary analysis for West Creek floodplain jams
In this code, I'm checking the data we collected at West Creek, calculating a few additional reach characteristics, and finding simple correlations between jams and measured valley characteristics.

## 1. Upload jam Characteristics
Both jam characteristics and reach characteristics (below) were copied into a .csv from the Google Sheets data file in the West Creek Google Drive. 
```{r}
#upload jam data from .csv
jam_data <- read.csv('E:/WestCreek/data/WC_jam_orig.csv', header = TRUE) %>%
  tibble::add_row(reach = 32) %>%
  mutate_at('woodvolume_m3', ~replace(., is.na(.), 0))
```

## 2. Upload reach characteristics and join the two datasets by reach number
This creates one master file that we can use to calculate future reach characteristics. 
```{r}
#upload reach data from .csv
reach_data <- read.csv('E:/WestCreek/data/WC_reach_orig.csv', 
                       header = TRUE)

#join reach and jam data by reach ID
master <- inner_join(jam_data, reach_data, by = 'reach')

```

## 3. Calculate more reach characteristics
Here, I'm going to calculate the total number of jams per reach, the number of organic jams per reach, the number of mixed/LW jams per reach, and the average jam size. 

```{r}
# count number of jams of each type per reach
count_jam_types <- master %>%
  #filter(!is.na(woodvolume_m3))%>%
  group_by(reach, jam_type)%>%
  summarise(numberofjams = n()) %>%
  ungroup() %>%
  spread(jam_type, numberofjams) %>%
  mutate_if(is.integer, ~replace(., is.na(.), 0)) %>%
  mutate(numberofjam_LWmixed = LW + mixed) %>%
  mutate(numberofjam_organic = organic) %>% 
  select(reach, numberofjam_LWmixed, numberofjam_organic)

#count total number of jams per reach
total_jams <- master %>%
  group_by(reach) %>%
  summarise(jam_total = n())

#edit reach with zero jams
total_jams$jam_total[total_jams$reach == 32] <- 0

#create reach df with all characteristics and add jam frequency
full_reach_ch <- master %>%
  mutate_at('woodvolume_m3', ~replace(., is.na(.), 0)) %>%
  group_by(reach) %>%
  summarise(jam_size_avg_m3 = mean(woodvolume_m3)) %>%
  inner_join(., count_jam_types, by = 'reach') %>%
  inner_join(., total_jams, by = 'reach') %>%
  inner_join(., reach_data, by = 'reach') %>%
  mutate(jam_per_m = jam_total/reach_length_m)

```

## 4. Normal distributions
In order to look at simple correlations, understanding whether the data are normal is important for determining correlation method. Here, I investigate normality using shapiro tests. Data are normal if p > 0.05.

### Normality of variables in jam dataset
```{r}
norm_test_jam <- jam_data %>%
  select_if(is.numeric) %>%
  map(shapiro.test)%>%
  map_df(pluck, 2) %>%
  gather(., variable, shapiro_p_value, reach:tot_blockage, 
         factor_key = TRUE)
pander(norm_test_jam, caption ='Shapiro-Wilk Test, Jam Data')
```

No variables in reach data are normal. Log transformation or sqrt transformation can be used to make data normal. 

\newpage
#### Log transformation of jam data
```{r}
log_test_jam <- jam_data %>%
  select_if(is.numeric) %>%
  select(-c(longitude)) %>%
  map_df(log10) %>%
  map(shapiro.test) %>%
  map_df(pluck, 2)%>%
  gather(., log_variable, shapiro_p_value, reach:tot_blockage, 
         factor_key = TRUE)
pander(log_test_jam, caption = 'Shapiro-Wilk Test, log jam data')
```

Log_10() transformation makes jam width and pin DBH (3) normal. 

\newpage
#### Transformation of jam response variables
The response variable of interest at the jam level is wood volume. Here, I'm testing if a square root, log10, or natural log transformation will make it normal. 
```{r}
shapiro.test(sqrt(jam_data$woodvolume_m3)) #not normal
shapiro.test(log10(jam_data$woodvolume_m3)) #not normal
shapiro.test(log(jam_data$woodvolume_m3)) #not normal
```

No transformation makes wood volume normal at the jam level. 

\newpage
### Normality of variables in reach dataset
```{r}
norm_test_reach <- full_reach_ch %>%
  select_if(is.numeric) %>%
  map(shapiro.test)%>% map_df(pluck, 2) %>%
  gather(., variable, shapiro_p_value, reach:jam_per_m, 
         factor_key = TRUE)
pander(norm_test_reach, caption = 'Shapiro-Wilk test, reach data')
```

Normally distributed variables: reach, total number of jams, average slope, average valley width, floodplain area, basal area, jam frequency. 

#### Transformation of reach response variables 
Total number of jams is normally distributed at the reach level, but total woodload and woodload per area are not. Here, I check to see if square root, log10, or natural log transformation will change that. 

```{r, eval = FALSE}
## testing for total woodload 
shapiro.test(sqrt(full_reach_ch$total_woodload_m3)) #normal

## testing for woodload per area
shapiro.test(sqrt(full_reach_ch$woodload_m3perm2)) #normal

```
Square root transformation makes the remaining two response variables normal. 

\newpage
## 5. Look at simple comparisons
Now that we have all of the dataframes and variables we need, we can perform some simple, univariate comparisons.

### Starting with jam level variables
Comparing wood volume per jam to other jam variables. The table below contain the correlation coefficient for different variables with regard to wood volume (the cor() function in R). 

The strongest correlation to wood volume (in m3) is DBH.

```{r, echo = F}
#create a df with variables for rcorr analysis
jam_cor <- select_if(jam_data, is.numeric) %>%
  dplyr::select(-c(2,3,4,5,6,7,11,14,16,17,19)) %>%
  mutate_if(is.numeric, ~replace(., is.na(.), 0))

#rcorr analysis to get r and p values
jam_rcorr <- rcorr(as.matrix(jam_cor), type = 'spearman')

#create table of r and p values just for wood volume
jam_table <- data.frame(jam_rcorr$r, jam_rcorr$P)%>%
  dplyr::select(woodvolume_m3, woodvolume_m3.1)%>%
  round(., 2) %>%
  mutate(var_name = colnames(jam_cor)) %>%
  rename(r_value = 1, p_value = 2, variable = 3)%>%
  mutate_if(is.numeric, ~replace(., is.na(.), 1))

#rearrange variables and get rid of r values that aren't significant
jam_table <- jam_table[,c(3,1,2)]
jam_table$r_value[jam_table$p_value > 0.05] <- " "

view(jam_table)

kable(jam_table, 'latex', booktabs = T, caption = 'Jam Correlation (R values if significant)') %>%
  add_header_above(c('','woodvolume_m3'=2))%>%
  kable_styling(latex_options = 'hold_position', full_width = F)

```


### Moving on to reach level variables
See page below for table of correlation values. Significant values are bolded. Average jam size and woodload per unit area did not significantly correlate to any reach scale characteristics. These two variables were correlated to each other, which is not an interesting relationship. 

Total woodload in m3 is signifcantly correlated to average slope and river left confinement (r > 0.5 and p<0.05). 

The number of jams (both organic and LW/mixed) is significantly correlated to valley bottom width, river left confinement, floodplain area, and basal area (r > 0.5 and p < 0.05). 

\newpage

```{r, echo = F}
#subset all variables to be used in correlation matrix
reach_cor<- select_if(full_reach_ch, is.numeric) %>%
  dplyr::select(-c(1,6,8,9,14,19)) %>%
  rename(Avg_Jam_Size = 1, Jams_LWmixed = 2, Jams_org = 3, Jams_total = 4,
         bf_width = 5, Avg_slope = 6, Valley_width= 7, RL_conf = 8,
         fp_area_m2 = 9,basal_tally = 10, basal_m2perha = 11, 
         tot_woodload = 12,woodload_m2 = 13)
  
#create correlation matrix
reach_rcorr <- rcorr(as.matrix(reach_cor), type = 'spearman')

emphasize.strong.cells(which(reach_rcorr[[3]]<0.05, arr.ind = TRUE))
pander(round(reach_rcorr[[1]], 2), caption = 'Reach Correlation Values (bolded if significant)')
```

\newpage
## 6. Looking at non-linear relationships
Some relationships between variables of interest and predictors may be non-linear. Looking at plots could be a good idea for revealing these relationships. 

### Jam scale plots of interest
We originally thought that there would be a predictable pattern in bankfull width and distance from channel to woodvolume per jam. However, plots reveal no real pattern in these data. 
```{r out.height = '25%', echo = FALSE}
## woodvolume vs. bf width
ggplot(jam_cor, aes(x = bf_width_m, y = woodvolume_m3, color = reach)) +
  geom_point() + 
  theme_few() + 
  labs(x = 'bankfull width (m)', y = 'wood volume (m3)', 
       title = 'woodvolume vs. bf width')

## wood volume vs. distance from channel
ggplot(jam_cor, aes(x = dist_from_channel_m, y = woodvolume_m3, 
                    color = reach)) + 
  geom_point() + 
  theme_few() + 
  labs(x = 'distance from ch (m)', y = 'wood volume (m3)', 
       title = 'woodvolume vs. distance from channel')
```

### Reach scale plots of interest
At the reach scale, we thought average jam size would scale non-linearly with basal area (largest jams at some intermediate basal area). Using log transformed average jam size, there appears to be the expected non-linear pattern. Additionally, I plotted jam frequency vs. basal area - which ended up being a linear function with a significant R2. 

```{r, echo = FALSE, out.height = '25%'}
## average jam size (log transformed) vs. basal area
ggplot(reach_cor, aes(x = basal_m2perha, y = Avg_Jam_Size)) + 
  geom_point(size = 2) + 
  theme_few() + 
  labs(x = 'basal area (m2/ha)', y = 'Average Jam Size', 
       title = 'basal area vs. average jam size')

## jam frequency vs. basal area 
ggplot(reach_cor, aes(x = basal_m2perha, y = jam_per_m)) + 
  geom_point(size = 2) + 
  theme_few() + 
  labs(x = 'basal area (m2/ha)', y = 'Jam Frequency', 
       title = 'basal area vs. jam freq')
```
\newpage

### Method for including non-linear relationship in regressions
Used in the models above for basal area or tally. A second degree polynomial explains the relationship between jams and basal area. This is not a significant correlation.
```{r, echo = FALSE, out.height = '25%'}
quad <- lm(Avg_Jam_Size~poly(basal_m2perha,2), data= reach_cor)
summary(quad)
#plot(quad)

qplot(basal_m2perha,Avg_Jam_Size, data = reach_cor) + 
  geom_smooth(method = 'lm', formula = y~poly(x,2), se = FALSE)
```
\newpage

## 7. Multiple regression models
Now that normality and correlation has been checked, multiple linear regressions for a few variables of interest can be created. Here, non-significant variables have been removed from the models. Transformed variables are used in all jam and reach scale models except for the 'Total Number of Jams' model (see last model in list). Untransformed variables produced a very significant model in this case, while no model with transformed variables was signficant.  

### Model for wood volume at the jam scale
Total blockage and number of pins are the only significant (p < 0.05) predictors according to the full model. Therefore, it makes statistical sense to dredge other predictors from the model. __The model has a very low R2 value.__ Also, residuals are not normal, no homoscedacity.

```{r}
## create dataframe excluding reaches without jams
jam_vol_df <- jam_data %>%
  filter(!reach==32) 

## full model for woodvolume at jam scale
jam_vol_mod_full <- lm(data = jam_vol_df, woodvolume_m3 ~ 
                         ht_above_bf_m +
                         dist_from_channel_m +
                         bf_width_m + 
                         obstruction_index +
                         pins+
                         tot_blockage)
#summary(jam_vol_mod_full)

## dredge full model
options(na.action = "na.fail")
jam_dredge <- dredge(jam_vol_mod_full, extra = c("R^2"))
head(jam_dredge)

```


\newpage

### Model for woodload per area
No significant model
```{r}
## full model for woodload per area
woodload_mod_full <- lm(sqrt(woodload_m3perm2) ~ bankfull_width_ave_m +
                          jam_total +
                          RL_confinement +
                          poly(basal_area_tally_ave,2), 
                          data = full_reach_ch)
#summary(woodload_mod_full)

## dredge full model
woodload_dredge <- dredge(woodload_mod_full, extra = c("R^2"))
head(woodload_dredge)

## dredged model with highest AICc is not significant
woodload_final <- lm(sqrt(woodload_m3perm2) ~ jam_total, data = full_reach_ch)
summary(woodload_final)
```


\newpage
### Model for total woodload
No significant model. 
```{r}
## full model for total woodload
tot_woodload_mod_full <- lm(sqrt(total_woodload_m3) ~ 
                              bankfull_width_ave_m +
                              jam_total +
                              RL_confinement + 
                              poly(basal_area_tally_ave,2), 
                            data = full_reach_ch)
#summary(tot_woodload_mod_full)

## dredge full model
tot_woodload_dredge <- dredge(tot_woodload_mod_full, extra = c("R^2"))
head(tot_woodload_dredge)


## dredged model with highest AICc is not significant
tot_woodload_final <- lm(sqrt(total_woodload_m3) ~ jam_total, 
                         data = full_reach_ch)
summary(tot_woodload_final)
#plot(tot_woodload_final)
```


\newpage
### Model for total number of jams
Bankfull width and RL_confinement are the only significant model variables. This is a significant model (r2 = 0.74, p <<0.01).
```{r}
## full model for number of jams
number_model_full <- lm(Jams_total ~ bf_width +
                          RL_conf +
                          poly(basal_tally, 2), data = reach_cor)
#summary(number_model_full)

## dredge full model
number_dredge <- dredge(number_model_full, extra = c("R^2"))
head(number_dredge)

## dredged model for number of jams per reach is significant!!
number_final <- lm(Jams_total ~ bf_width + RL_conf, 
                         data = reach_cor)
summary(number_final)
```

## 11. Conclusions
While there is a weak non-linear relationship between jam characteristics and basal area, the strongest modeled relationship was between total number of jams and bankfull width + RL confinment. Geometric characteristics are more significant than forest characteristics in multiple regressions. 





